
```{r setup-02, include=FALSE}
library(tidyverse)
library(arrow)
library(viridis)
library(patchwork)
library(ggridges)
library(scales)

df <- read_parquet("../data/processed/articles_with_sentiment.parquet") |>
  mutate(
    published_date = as.Date(published_date),
    bert_net = bert_positive - bert_negative
  )

# Country-level averages
country_avg <- df |>
  group_by(country) |>
  summarise(
    positive = mean(bert_positive, na.rm = TRUE),
    neutral  = mean(bert_neutral,  na.rm = TRUE),
    negative = mean(bert_negative, na.rm = TRUE),
    net      = mean(bert_net,      na.rm = TRUE),
    gdelt    = mean(gdelt_tone,    na.rm = TRUE),
    n        = n(),
    .groups  = "drop"
  )

# Outlet-level averages
outlet_avg <- df |>
  group_by(country, outlet_name) |>
  summarise(
    net = mean(bert_net, na.rm = TRUE),
    n   = n(),
    .groups = "drop"
  )
```

## Sentiment Heatmap

Country × sentiment dimension (avg BERT probabilities).

```{r heatmap, fig.height=5, fig.width=8}
country_avg |>
  pivot_longer(c(positive, neutral, negative), names_to = "sentiment", values_to = "prob") |>
  mutate(sentiment = factor(sentiment, levels = c("positive", "neutral", "negative"))) |>
  ggplot(aes(x = sentiment, y = reorder(country, net), fill = prob)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(prob, 2)), size = 3.5, color = "white") +
  scale_fill_viridis(option = "plasma", name = "Avg probability") +
  labs(
    title = "Average BERT sentiment probabilities by country",
    x = "Sentiment class", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", legend.key.width = unit(2, "cm"))
```

## Net Sentiment by Outlet

```{r outlet-bars, fig.height=9, fig.width=10}
corpus_mean <- mean(df$bert_net, na.rm = TRUE)

outlet_avg |>
  mutate(
    direction = ifelse(net > corpus_mean, "above average", "below average"),
    outlet_name = reorder(outlet_name, net)
  ) |>
  ggplot(aes(x = net, y = outlet_name, fill = direction)) +
  geom_col() +
  geom_vline(xintercept = corpus_mean, linetype = "dashed", color = "grey40") +
  facet_wrap(~country, scales = "free_y", ncol = 3) +
  scale_fill_manual(
    values = c("above average" = "#2196F3", "below average" = "#F44336"),
    guide = "none"
  ) +
  labs(
    title = "Net sentiment (positive − negative) by outlet",
    subtitle = "Dashed line = corpus mean",
    x = "Net sentiment score", y = NULL
  ) +
  theme_minimal(base_size = 11)
```

## Sentiment Distribution (Ridge Plot)

```{r ridges, fig.height=7, fig.width=9}
df |>
  mutate(country = reorder(country, bert_net, FUN = mean, na.rm = TRUE)) |>
  ggplot(aes(x = bert_net, y = country, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +
  scale_fill_viridis(option = "cividis", name = "Net sentiment") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "white", linewidth = 0.8) +
  labs(
    title = "Distribution of net sentiment (positive − negative) by country",
    x = "Net sentiment score", y = NULL
  ) +
  theme_minimal(base_size = 13)
```

## Diverging Bar: Most vs Least Positive Outlets

```{r diverging, fig.height=7, fig.width=9}
outlet_avg |>
  mutate(
    label = paste0(outlet_name, " (", country, ")"),
    label = reorder(label, net)
  ) |>
  ggplot(aes(x = net, y = label, fill = net > 0)) +
  geom_col() +
  geom_vline(xintercept = 0, color = "grey30") +
  scale_fill_manual(
    values = c("TRUE" = "#1976D2", "FALSE" = "#D32F2F"),
    labels = c("Net positive", "Net negative"),
    name   = NULL
  ) +
  labs(
    title = "Outlets ranked by net sentiment",
    x = "Net sentiment (positive − negative)", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

## Model Confidence by Country

GDELT tone scores are not available via the Doc API (they are only computed in the BigQuery GKG pipeline), so cross-validation against GDELT is not possible here. As an alternative quality check, the chart below shows the distribution of BERT **prediction confidence** — the probability assigned to the winning class — for each country. High median confidence (> 0.8) indicates the model is making clear, unambiguous predictions rather than borderline calls.

```{r validation, fig.height=6, fig.width=9}
df |>
  mutate(
    confidence = pmax(bert_positive, bert_neutral, bert_negative, na.rm = TRUE),
    country    = reorder(country, confidence, FUN = median, na.rm = TRUE)
  ) |>
  ggplot(aes(x = confidence, y = country, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 1.8, rel_min_height = 0.01, quantile_lines = TRUE, quantiles = 2) +
  scale_fill_viridis(option = "magma", name = "Confidence", direction = -1) +
  scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0.3, 1)) +
  geom_vline(xintercept = 0.8, linetype = "dashed", color = "white", linewidth = 0.6) +
  labs(
    title = "BERT prediction confidence by country",
    subtitle = "Vertical line = 0.80 threshold · median marked by quantile line",
    x = "Confidence (probability of predicted class)", y = NULL
  ) +
  theme_minimal(base_size = 13)
```
